---
title: "Simulating birth-death-immigration processes"
author: "A. J. Rominger"
date: "10/4/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'm working on figuring this out.  Rough ideas right now about how to do it.  Below are three different attempts, the first two seemingly wrong mathematically, the second taken from Bailey, 1964 and Bolker, 2007, thus seemingly seems to be mathematically correct, but yeilds results that I'm not sure about (lots of immigrants very quickly).

## Attempt one

Assumes each event in each time step is competed for by all $n_J$ species

```{r, fig.width=4, fig.height=4, fig.align='center'}
library(pika)

## the metacommunity
b <- 0.01
nJ <- 500
# J <- rfish(nJ, 0.01)
J <- rep(1, nJ)
J <- J / sum(J)

## the local community
nTime <- 50000

## initialization
x <- matrix(0, nrow = nJ, ncol = nTime)
x[sample(1:nJ, 1, prob = J), 1] <- 1

## rates: mu > lambda + nu
mu <- 1
lambda <- 1
nu <- 0.1

## the simulation
sim <- lapply(2:nTime, function(i) {
    thisX <- x[, i-1]
    
    event <- sample(1:3, 1, prob = c(c(mu, lambda) * sum(thisX), nu))
    if(event == 1) {
        spID <- sample(1:nJ, 1, prob = thisX)
        thisX[spID] <- thisX[spID] - 1
    } else if(event == 2) {
        spID <- sample(1:nJ, 1, prob = thisX)
        thisX[spID] <- thisX[spID] + 1
    } else {
        spID <- sample(1:nJ, 1, prob = J)
        thisX[spID] <- thisX[spID] + 1
    }
    
    x[, i] <<- thisX
    
    return(NULL)
})


foo <- rowSums(x)
foo <- foo[foo > 0]
plot(sort(foo, TRUE), log = 'y')

i <- order(rowSums(x), decreasing = TRUE)
plot(x[i[4], ], type = 'l')
plot(colSums(x), type = 'l')
```

## Attempt two

Assumes all species have independent dynamics, but uses a poor approximation for combining immigration process with birth-death.

I'm not running the below code because it takes a while, but the issue is that the local community goes completely extinct for the majority of the time.

```{r, eval=FALSE}
library(pika)

## helper function to take time and pop size and return pop size at fixed intervals
interp <- function(tt, x, slices) {
    newx <- sapply(split(x, cut(tt, slices)), function(xi) {
        out <- xi[length(xi)]
        if(length(out) == 0) out <- NA
        
        return(out)
    })
    
    names(newx) <- NULL
    
    i <- which(!is.na(newx))
    if(is.na(newx[1])) i <- c(1, i)
    
    ## fill in missing with last non-na value
    out <- rep(newx[i], times = diff(c(i, length(newx) + 1)))
    out[is.na(out)] <- 0
    
    return(out)
}


## the metacommunity
b <- 0.01
nJ <- 100
J <- rfish(nJ, b)
J <- nJ * J / sum(J)


## initialization
x <- numeric(nJ)
# x[sample(1:nJ, 1, prob = J)] <- 1
x[1] <- 1
maxTime <- 5000
timeSlice <- seq(-100, maxTime, 100)

## rates. note: nu is imm rate of avg spp in metacomm, because average value in J is 1
mu <- 1
lambda <- 1
nu <- 0.1

sim <- lapply(1:nJ, function(i) {
    nImm <- rpois(1, nu * J[i] * maxTime)
    if(x[i] > 0) nImm <- nImm + 1
    
    if(nImm > 0) {
        imm <- runif(nImm, 0, maxTime) # immigration times
        if(x[i] > 0) imm[1] <- 0
        
        out <- lapply(imm, function(immTime) {
            thisX <- cumsum(c(1, sample(c(-1, 1), round(maxTime * (mu + lambda) * 100), 
                                        replace = TRUE, prob = c(mu, lambda))))
            if(!any(thisX == 0)) {
                ext <- sample(length(thisX), 1, prob = 1/thisX)
                thisX[ext] <- 0
            }
            
            thisX <- thisX[1:which(thisX == 0)[1]]
            thisT <- cumsum(c(immTime, rexp(length(thisX) - 1, 
                                            (lambda + mu) * thisX[-length(thisX)])))
            
            return(interp(thisT, thisX, timeSlice))
        })
        
        out <- rowSums(do.call(cbind, out))
    } else {
        out <- rep(0, length(timeSlice) - 1)
    }
    
    return(out)
})

sim <- do.call(rbind, sim)

par(mar = c(3, 3, 0, 0) + 0.5, mgp = c(1.75, 0.5, 0))
plot(colSums(sim), type = 'l', xlab = 'Iterations', ylab = 'Total community abundance')
```


## Attempt three

Just a negative binomial sampling process.

```{r, fig.width=4, fig.height=4, fig.align='center'}
library(pika)
library(viridis)
library(socorro)

## function to simulate birth-death-immigration process starting at X(t = 0) = 0
rbdi <- function(n, lambda, mu, nu, t) {
    rnbinom(n, size = nu/lambda, mu = nu / (lambda - mu) * (exp((lambda - mu) * t) - 1))
}

## metacommunity parameters
b <- 0.01
nJ <- 500
J <- rfish(nJ, b)
# J <- rep(1, nJ)
J <- nJ * J / sum(J)

## local community parameters
lambda <- 1
mu <- 1.0001
nu <- 0.05 # imm for avg spp in metacomm

## times to simulate
tt <- exp(seq(log(10), log(200000), length.out = 7))


sim <- lapply(tt, function(ti) {
    out <- replicate(500, sort(rbdi(nJ, lambda, mu, nu * J, ti), decreasing = TRUE))
    return(cbind(m = rowMeans(out), t(apply(out, 1, quantile, probs = c(0.025, 0.975)))))
})

par(mar = c(3, 3, 0, 0) + 0.5, mgp = c(1.75, 0.5, 0))
plot(sim[[1]][sim[[1]][, 1] > 0, 1], log = 'y', xlim = c(0, 155), ylim = c(1e-03, 50000), 
     col = viridis(length(tt))[1], yaxt = 'n', 
     xlab = 'Rank', ylab = 'Mean abundance')
for(i in 2:length(tt)) points(sim[[i]][sim[[i]][, 1] > 0, 1], col = viridis(length(tt))[i])
logAxis(2, expLab = TRUE)
```

The time evolution of these SADs goes from early (purple) to late (yellow).
