---
title: "Troubleshooting the Psi distribution"
author: "Andrew Rominger"
date: "4/24/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The current implementation of the $\Psi$ distribution in *meteR* doesn't work well for large values of the state variables. For example for BCI:

```{r, cache=TRUE}
library(meteR)
bci <- read.csv('~/Dropbox/Research/data/stri/BCIS.csv', as.is = TRUE)
bci <- bci[bci$year == 1995, ]

thisESF <- meteESF(bci$spp, bci$count, bci$dbh^2)
thisIPD <- ipd(thisESF)
plot(thisIPD, ptype = 'rad', log = 'xy', ylim = c(1, 10^6))
```

We could see how well the approximation for the rank $\Psi$ distribution (John's eq 7.37 in the book) works for such large values:

```{r, cache=TRUE}
rankPsi <- function(La, N0, r) {
  la1 <- La[1]
  la2 <- La[2]
  b <- la1 + la2
  
  f <- 1/la2 * log((b * N0 + r - 0.5) / (r - 0.5)) - la1 / la2
  
  return(f)
}

plot(thisIPD, ptype = 'rad', log = 'xy', ylim = c(1, 10^6), add.legend = FALSE)
lines(1:thisESF$state.var['N0'], rankPsi(thisESF$La, thisESF$state.var['N0'], 1:thisESF$state.var['N0']), 
      col = 'blue')
legend('topright', legend = c('data', 'meteR', 'rank approx'), col = c('black', 'red', 'blue'), 
       pch = c(1, NA, NA), lty = c(NA, 1, 1), bty = 'n')
```

Yup, looks a lot better.  But how does it look for smaller values?

```{r, cache=TRUE}
arthESF <- meteESF(arth$spp, arth$count, arth$mass^0.75)
arthIPD <- ipd(arthESF)
plot(arthIPD, ptype = 'rad', log = 'xy', ylim = c(1, 1000), add.legend = FALSE)
lines(1:arthESF$state.var['N0'], rankPsi(arthESF$La, arthESF$state.var['N0'], 1:arthESF$state.var['N0']), 
      col = 'blue')
legend('topright', legend = c('data', 'meteR', 'rank approx'), col = c('black', 'red', 'blue'), 
       pch = c(1, NA, NA), lty = c(NA, 1, 1), bty = 'n')
```

Still ok there.  How about even smaller:

```{r, cache=TRUE}
randESF <- meteESF(S0 = 50, N0 = 100, E0 = 1000)
randIPD <- ipd(randESF)
randIPD$data <- sort(randIPD$r(100), decreasing = TRUE)

plot(randIPD, ptype = 'rad', log = 'y', add.legend = FALSE)
lines(1:randESF$state.var['N0'], rankPsi(randESF$La, randESF$state.var['N0'], 1:randESF$state.var['N0']), 
      col = 'blue')
legend('topright', legend = c('data', 'meteR', 'rank approx'), col = c('black', 'red', 'blue'), 
       pch = c(1, NA, NA), lty = c(NA, 1, 1), bty = 'n')
```

There the approximation deviates. So the next step is to incorperate the approximation into the *meteR* framework for only big values of the state variables. To do that we need to modify the `rankPsi` function and use it to create a quantile function.

To start we remove the `0.5` modifier on rank `r`:
```{r}
rankPsi <- function(La, N0, r) {
  la1 <- La[1]
  la2 <- La[2]
  b <- la1 + la2
  
  f <- 1/la2 * log((b * N0 + r) / (r)) - la1 / la2
  
  return(f)
}
```

The rank argument `r` of this new `rankPsi` function now has a one-to-one correspondance with probability increments of $1 - \frac{r}{N_0}$ for $r \in \{0 \ldots N_0\}$, although the boundary values 0 and $N_0$ are problematic and will require further care.

Now instead of having `r` as the input, we want `p` (i.e. the probability) as the input. We just solve for $r$ in  
$$
p = 1 - \frac{r}{N_0}
$$
$$
r = N_0(1 - p)
$$



And now we can make our quantile function:
```{r}
qPsi <- function(p, La, N0, E0, log.p = FALSE) { # we need E0 to set the upper bound
  la1 <- La[1]
  la2 <- La[2]
  b <- la1 + la2
  
  if(log.p) p <- exp(p)
  r <- (1 - p) * N0
  
  f <- 1/la2 * log((b * N0 + r) / (r)) - la1 / la2
  
  # take care of boundary conditions
  f[p == 1] <- E0
  f[p == 0] <- 1
  
  return(f)
}
```

Now we just have to decide where to draw the cutoff between the numerical solution (for smaller state variables) and the analytical solution (for large values):

```{r, cache=TRUE}
library(socorro)
library(viridis)

# set a range of state variables to test
S <- 2^(6:9)
Nfact <- 2^(1:6)
Efact <- 2^(1:6)

stateVar <- expand.grid(S0 = S, N0 = Nfact, E0 = Efact)
stateVar$N0 <- stateVar$N0 * stateVar$S0
stateVar$E0 <- stateVar$E0 * stateVar$N0

# loop over state variables to calculate MSE between current quantile function
# and analytical approximation
foo <- lapply(1:nrow(stateVar), function(i) {
  thisESF <- meteESF(S0 = stateVar$S0[i], N0 = stateVar$N0[i], E0 = stateVar$E0[i])
  thisIPD <- ipd(thisESF)
  
  q1 <- thisIPD$q(seq(0.001, 0.999, length.out = 30))
  qApprox <- qPsi(seq(0.001, 0.999, length.out = 30), La = thisESF$La, 
                  N0 = thisESF$state.var['N0'], E0 = thisESF$state.var['E0'])
  
  qmse <- mean((q1 - qApprox)^2)
  
  return(c(thisESF$La, thisESF$state.var, mse = qmse))
})
foo <- do.call(rbind, foo)

# plot the results to determine where to use approximation
layout(matrix(1:2, nrow = 1), widths = 2:1)
par(mar = c(3, 3, 0, 0) + 0.5, mgp = c(2, 0.5, 0))

col <- quantCol(foo[, 'mse'], viridis(50), trans = 'log')
plot(foo[, 'N0'], foo[, 'E0'], col = col, 
     log = 'xy',
     axes = FALSE, frame.plot = TRUE, 
     xlab = expression(N[0]), ylab = expression(E[0]))
logAxis(1, expLab = TRUE)
logAxis(2, expLab = TRUE)

curve(1.4*10^5 - 200 * x^0.7, add = TRUE, from = 10^3, to = 10^4)
segments(x0 = 10^4, y0 = 1, y1 = 1.4*10^5 - 200 * (10^4)^0.7)
segments(x0 = 1, x1 = 10^3, y0 = 1.4*10^5 - 200 * (10^3)^0.7)

plot(sort(foo[, 'mse']), col = col[order(foo[, 'mse'])], 
     log = 'y', axes = FALSE, frame.plot = TRUE, 
     xlab = '', ylab = 'MSE')
logAxis(2, expLab = TRUE)
```

We'll use the numeric solution for state variable values below the funky curve.  We can make a function to return a boolean determining which region we're in:

```{r}
.psiNumeric <- function(N0, E0) {
  return(E0 > 1.4*10^5 - 200 * N0^0.7 |
    N0 > 10^4)
}
```

We'll incorperate that in the guts of the `meteR::ipd.meteESF` function like this:

```{r, eval=FALSE}
ipd.meteESF <- function(x,...) {
    if(is.na(x$state.var[3])) {
      stop('must provide metabolic rate data or E0 to calculate power distributions')
    }
  
    dat <- x$data$e
    if(is.null(dat)) {
        X <- NULL
    } else {
        X <- sort(dat, decreasing=TRUE)
    }
    
    this.eq <- function(epsilon, log=FALSE) {
        out <- metePsi(epsilon, la1=x$La[1], 
                       la2=x$La[2], Z=x$Z,
                       S0=x$state.var[1], 
                       N0=x$state.var[2], 
                       E0=x$state.var[3])
        
        if(log) out <- log(out)
        
        return(out)
    }

    this.p.eq <- function(epsilon, lower.tail=TRUE, log.p=FALSE) {
      b <- sum(x$La)
      la1 <- x$La[1]
      la2 <- x$La[2]
      Z <- x$Z
      S0 <- x$state.var[1]
      N0 <- x$state.var[2]
      
      out <- S0/(Z*N0*la2) * ((exp(-(N0-1) * (la1 + la2*epsilon)) - 1) / 
                                (exp(la1 + la2*epsilon) - 1) - 
                                (exp(-(N0-1)*b) - 1)/(exp(b) - 1))
      
      if(!lower.tail) out <- 1 - out
      if(log.p) out <- log(out)
      
      return(out)
    }
    
    this.q.eq <- ifelse(.psiNumeric(x$state.var['N0'], x$state.var['E0']), 
                        NULL, 
                        function(p, lower.tail = TRUE, log.p = FALSE) {
                          if(log.p) p <- exp(p)
                          if(!lower.tail) p <- 1 - p
                          
                          return(.qPsi(p, x$La, x$state.var['N0'], x$state.var['E0'], 
                                      log.p = FALSE))
                        })
    
    FUN <- distr::AbscontDistribution(d=this.eq, p=this.p.eq, q = this.q.eq,
                                      low1=1, low=1, up=x$state.var[3], up1=x$state.var[3],
                                      withgaps=FALSE,
                                      ngrid=distr::getdistrOption(
                                        'DefaultNrGridPoints')*10^2)
    
    out <- list(type='ipd', data=X, 
                d=this.eq, p=FUN@p, q=FUN@q, r=FUN@r,
                state.var=x$state.var, La=x$La)
    class(out) <- c('ipd', 'meteDist')
    
    return(out)
}
```
